---
title: "Description of `tobalciomodel` Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Description of Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(tobalciomodel)
```

## Input-Output Tables

The input-output table is that used by [Connolly et al. (2018)](https://strathprints.strath.ac.uk/63723/) (FAI hereafter). It used a table containing 106 sectors, 3 of which are alcohol specific sectors disaggregated from the larger Standard Industrial Classification (SIC) industries they belong to. [The Dataset produced for the study has been made available publicly here.](https://pureportal.strath.ac.uk/en/datasets/2010-uk-alcohol-consumption-disaggregated-ixi-table) and is incorporated into the `tobalciomodel` package.

The dataset includes the flow table (which describes intermediate consumption between industries) as well as other demand and production vectors needed for the IO model to function. Below shows a subset of the IO flow table - the first four sectors. 

```{r iotable}
tobalciomodel::iotable[1:4,2:5]
```

All purchases are measured in £m. Columns show purchases. Sector 3 purchases £0.775m of sector 1 output, £0.272m of sector 2 output, __£27.029m of its own output__, and £0.061m of sector 4 output. Conversely, rows show sales. Sector 2 makes no sales to sector 1, £262m sales to itself, £0.272m sales to sector 3 (*the other interpretation of the same figure. Purchases to sector 3, sales to sector 2*), and £0.986m sales to sector 4. 

Other variables include demand from households, final demand (the sum of all demand for goods for consumption, from households, governments, exports), household output (total employee compensation), total output (intermediate goods, plus imports, plus value added), total demand (intermediate demand plus final demand), and gross value added variables - total, and split into employee compensation, gross operating surplus, taxes.  

```{r names_iotable}
names(tobalciomodel::iotable[,c(1,108:116)])
```


## Alcohol Data

This section describes and summarises the Scottish and English/Welsh alcohol consumption data that is included in the package. The data were originally compiled by MESAS from Nielsen retail sales data and were used by FAI and Connolly et al. (2019) in their study of the wider economic impacts of alcohol policy. 

The variables included in the data are:

```{r mesas_names,echo=FALSE}
names(tobalciomodel::alcohol_data)
```

## Tobacco Data


## Elasticities

To model the change in demand which occurs due to an induced price change (through an increase in alcohol duty, or a minimum unit price) requires price elasticity of demand data for alcohol products. Elasticities for beer, cider, wine, spirits, and RTDs separately by on and off trade are obtained from Meng et al. (2014) and Collis et al. (2010). 

These elasticities are coded into `tobalciomodel` as package data in `tobalciomodel::elasticities_alc`:

```{r elasticities_names}
tobalciomodel::elasticities_alc
```

## Employment

Employment data is not typically included with IO tables, so in order to model the impacts of policy on employment this information must be obtained externally and matched to the sectors in the IO table. 

### Obtaining and Cleaning Employment Data

Employment data is sourced from the quarterly Labour Force Survey (LFS) and processed using `lfsclean` version 0.6.2. Employment data coded according to SIC-2007 industries has been available since 2008. We create annual employment figures by 4-digit industry to then map to the 106 industries used in the FAI input-output table. Employment is measured as total number of full-time equivalents (FTEs), with a part-time employee counted as 0.5.  

Firstly, we total the number of employees (excluding self-employed) by 4-digit industry within each quarterly dataset, using the LFS person weights to aggregate to the employment count into an estimate for the whole population. To obtain annual figures, we then average employment in each industry across the four quarters. Employment in 2020 is currently based only on quarter 1 data. This produces a panel of 4-digit industry employment over the period 2010-2020.  

### Matching Cleaned Employment Data to the FAI IO Table Sectors

An issue with the FAI table is that the industry codes do not align with the industry names in all cases. For example, IOC 94 is labelled as "Other Personal Service Activities" whereas SIC-2007 2-digit code 94 in fact relates to "Activities membership organisations" and "Other Personal Service Activities" is actually SIC-2007 division 96. We make the assumption that the data in the FAI table is correctly matched to industry titles, rather than the SIC codes, and accordingly match on the basis of titles rather than codes. 

In this way we match the SIC-2007 4-digit industries onto the 103 industries in the IO table used by FAI. In order to allocate employment in the sectors which are disaggregated into alcohol and non-alcohol, we use the output data in the IO table and assume employment is split according to the same proportion as output. e.g. if output in Wholesale and Retail Trade is split 80/20 between the non-alcohol and alcohol sectors, this same split is applied to employment. 

This produces employment figures for the 106 sectors in the FAI IO table and is found in the dataset `employment`, sorted in order of sector in the FAI tables and so can be easily matched to `iotable` on a one-to-one basis. Both headcount and FTE measures of employment are produced for each year 2010-2020 inclusive.  

```{r employment_names}
names(tobalciomodel::employment)
```

## References

- Collis, J., Grayson, A. and Johal, S. (2010) ‘Econometric Analysis of Alcohol Consumption in the UK Econometric Analysis of Alcohol Consumption in the UK’, HRMC working paper, 10, pp. 1–70. Available at: (https://www.gov.uk/government/publications/econometric-analysis-of-alcohol-consumption-in-the-uk).

- Connolly, K. (Creator) (18 Mar 2019). 2010 UK alcohol consumption disaggregated IxI table. University of Strathclyde. 2010_UK_Alcohol_consumption_disaggregated_IxI(.xlsx). 10.15129/cfb2d861-2856-437b-bd45-ed37dbee9900 available at: (https://pureportal.strath.ac.uk/en/datasets/2010-uk-alcohol-consumption-disaggregated-ixi-table) __[accessed 29/07/2020]__

- Connolly, K., Lisenkova, K. and McGregor, P., (2018). The Economic Impact of Changes in Alcohol Consumption in the UK. University of Strathclyde.  available at: (https://pure.strath.ac.uk/ws/portalfiles/portal/80224583/FAI_2018_The_economic_impact_of_changes_in_alcohol_consumption.pdf) __[accessed 29/07/2020]__

- Meng, Y., Brennan, A., Purshouse, R., Hill-McManus, D., Angus, C., Holmes, J., Meier, P.S (2014) ‘Estimation of own and cross price elasticities of alcohol demand in the UK-A pseudo-panel approach using the Living Costs and Food Survey 2001-2009’, Journal of Health Economics. Elsevier B.V., 34(1), pp. 96–103.
