---
title: "Input-Output Modelling using tobalciomodel"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Input-Output Modelling using tobalciomodel}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

suppressPackageStartupMessages(library(DiagrammeR))
suppressPackageStartupMessages(library(here))
suppressPackageStartupMessages(library(data.table))
suppressPackageStartupMessages(library(stapmr))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(ggthemes))
suppressPackageStartupMessages(library(pricepol))
suppressPackageStartupMessages(library(readxl))
suppressPackageStartupMessages(library(flextable))
suppressPackageStartupMessages(library(magrittr))
suppressPackageStartupMessages(library(knitr))
suppressPackageStartupMessages(library(tobalciomodel))

```

## Introduction

The purpose of this vignette is to describe how the `tobalciomodel` produces input-output analyses of the impact of changes in alcohol and tobacco consumption on the economy. The package can produce policy analysis by using the output of a price policy modelling simulation in the Sheffield Tobacco and Alcohol Policy Model (STAPM). Alternatively, using data stored in the package itself, simple "percentage change" scenarios can be modelled. 

To illustrate the use of the `pricepol` functions for different tasks, we have created example workflows. You can use these examples to help understand how the code works. At the moment, these workflows are accessible only to the project team. To access them, you will need to be added to the SPECTRUM WP4 Gitlab organisation.  

The workflows are listed below:  

- [Input output modelling](https://gitlab.com/SPECTRUM_Sheffield/projects/input-output-modelling)

The workflow for the input-output modelling to be described in this vignette is summarised in the figure below. 

```{r flowdiag, echo = F, fig.cap = "Input-output modelling workflow."}

DiagrammeR::grViz("digraph {

graph [layout = dot]

# define the global styles of the nodes. We can override these in box if we wish

node [shape = rectangle, style = filled, fillcolor = Linen]

a [label =  'STAPM Output \n EconCalc()', shape = folder, fillcolor = green]

subgraph cluster_0 {

graph[shape = rectangle, style = rounded, bgcolor = cyan]
label = 'tobalciomodel functions and data'

data1 [label = 'Alcohol & Tobacco \n Expenditure Data', shape = folder, fillcolor = green]
data1a [label = 'MESAS Report 2021', shape = folder, fillcolor = LemonChiffon]
data1b [label = 'Tobacco Bulletin', shape = folder, fillcolor = LemonChiffon]
data1c [label = 'Alcohol Bulletin', shape = folder, fillcolor = LemonChiffon]
data2 [label = 'Input-Output \n Table', shape = folder, fillcolor = green]
data3 [label = 'Employment', shape = folder, fillcolor = green]
data4 [label = 'Earnings', shape = folder, fillcolor = green]
data5 [label = 'Income Tax / NI', shape = folder, fillcolor = green]

data3a [label = 'Labour Force Survey', shape = folder, fillcolor = LemonChiffon]
data4a [label = 'Annual Survey of \n Hours and Earnings', shape = folder, fillcolor = LemonChiffon]


b [label =  'Read Inputs']
c [label =  'Generate Inputs']
d [label =  'Prepare \n Final Demand Vectors']
e [label =  'Read in the \n Input-Output table']
f [label =  'Calculate Multipliers & \n Leontief Inverse matrices']
g [label =  'Calculate Economic Effects']
h [label =  'Process Outputs']
}

out1 [label=   'Direct and Indirect \n Economic Impacts', shape = oval, fillcolor = gold]
out2 [label=   'Sector Specific \n Economic Impacts', shape = oval, fillcolor = gold]

# edge definitions with the node IDs
{b c} -> d -> e -> f -> g -> h -> {out1 out2} 
                  a -> b
                  data1 -> c
                  data1a -> data1
                  data1b -> data1
                  data1c -> data1
                  data2 -> {d e}
                  data3a -> data3
                  data3 -> g
                  data4a -> data4
                  data4 -> g
                  data5 -> g

}", width = 768, height = 500)

```

## Data 

A number of external data sources are required to operationalise the input-output methodology. The most important component is the input-output table - a national statistic recording all inter-industry purchases and sales as well as total industry output, employee compensation, and household/government demands from each sector. Aggregate data on alcohol and tobacco expenditures are required as inputs to the model. 

### Input-Output Tables

The input-output table is that used by [Connolly et al. (2018)](https://strathprints.strath.ac.uk/63723/) (FAI hereafter). It used a table containing 106 sectors, 3 of which are alcohol specific sectors disaggregated from the larger Standard Industrial Classification (SIC) industries they belong to. [The Dataset produced for the study has been made available publicly here.](https://pureportal.strath.ac.uk/en/datasets/2010-uk-alcohol-consumption-disaggregated-ixi-table) and is incorporated into the `tobalciomodel` package.

The dataset includes the flow table (which describes intermediate consumption between industries) as well as other demand and production vectors needed for the IO model to function. Below shows a subset of the IO flow table - the first four sectors. 

All purchases are measured in £m. Columns show purchases. Sector 3 purchases £0.775m of sector 1 output, £0.272m of sector 2 output, __£27.029m of its own output__, and £0.061m of sector 4 output. Conversely, rows show sales. Sector 2 makes no sales to sector 1, £262m sales to itself, £0.272m sales to sector 3 (*the other interpretation of the same figure. Purchases to sector 3, sales to sector 2*), and £0.986m sales to sector 4. 

Other variables include demand from households, final demand (the sum of all demand for goods for consumption, from households, governments, exports), household output (total employee compensation), total output (intermediate goods, plus imports, plus value added), total demand (intermediate demand plus final demand), and gross value added variables - total, and split into employee compensation, gross operating surplus, taxes.

### Price, Consumption, and Expenditure Data

Aggregate consumption, price, and expenditures data are required to provide the inputs the model. As the input-output tables report expenditures in basic prices (market prices net of tax), the inputs to the model must also be expressed in basic prices. For both tobacco and alcohol products (two tobacco and four alcohol, divided into on and off-trade) a time series of expenditures at basic prices is calculated for 2010-2020. 

#### Alcohol

The data used to compile the originally compiled by the Public Health Scotland programme Monitoring and Evaluating Scotlands Alcohol Strategy (MESAS) from Nielsen retail sales data and were used by FAI and Connolly et al. (2019) in their study of the wider economic impacts of alcohol policy. We use the data from the [MESAS monitoring report 2021](https://www.publichealthscotland.scot/publications/mesas-monitoring-report-2021/) for information on aggregate consumption. These data contain mean prices, units of alcohol consumed per head, and population counts reported by year, alcohol product, and on vs off trade. Data are presented separately for Scotland and for England & Wales.

In processing the data we construct total units of alcohol consumed by multiplying units per adult by total adult population. England & Wales data is combined with Scotland data by summing up consumption across the nations and calculating a consumption-weighted mean price. We collapse the alcohol categories to beer, cider, wine, and spirits.

As the modelling is at the level of the UK, we adjust consumption figures accordingly before calculating total expenditure. To do this we assume that consumption patterns in Northern Ireland are similar to those for Great Britain and apply a simple upshift. Using [national population projections](https://www.nomisweb.co.uk/query/construct/summary.asp?mode=construct&version=0&dataset=2009) data, we calculate that the population of Northern Ireland is 2.84% of the population of Great Britain in each year from 2018 to 2030. Total units of alcohol consumed in MESAS data are therefore upshifted by a multiplier of 1.0284 to produce a UK consumption estimate. To calculate total UK expenditures at market prices, we assume the mean prices calculated for Great Britain to be representative of the UK mean price.  

In order to calculate expenditure in basic prices, we merge in alcohol duty receipts data by calendar year from the [May - July 2021 Alcohol Bulletin tables](https://www.gov.uk/government/statistics/alcohol-bulletin). These data are presented for the four different alcohol duty regimes administered by HMRC; beer, cider, wine, and spirits (including RTDs). The duty receipts data makes no distinction between on and trade purchases, so the MESAS consumption data are used to apportion the total duty receipts for each product into on-trade and off-trade. The adjusted duty total figures are then subtracted from the upshifted expenditure at market prices along with VAT to give expenditure at basic prices. 

#### Tobacco

A similar approach is taken to the construction of aggregate tobacco expenditures data, using primarily the duty and clearances data in the [July 2021 Tobacco Bulletin](https://www.gov.uk/government/statistics/tobacco-bulletin). Price data are also obtained from the ONS for a 20-pack of cigarettes, and the consumer price index (CPI) for tobacco products is used to estimate the time series of roll-your-own tobacco prices (per 100g) from the estimated price in 2020 calculated from an average of prices taken from supermarket websites in December 2020. 

Clearance data for factory-made (reported in millions of sticks) and roll-your-own (reported in thousands of kilograms) are converted, respectively, into packs of 20 sticks and packs of 100g. These are then multiplied by the prices to obtain estimated total expenditure at market prices. As with alcohol, VAT is calculated from this figure and subtracted, along with total duty receipts, to give total expendture at basic prices.  

### LFS Employment Data

Employment and earnings data are not included with IO tables, so in order to model the impacts of policy on employment this information must be obtained externally and matched to the sectors IO table. Employment data is obtained from the quarterly Labour Force Survey (LFS) and processed using the `lfsclean` version 0.6.2 package. Employment data coded according to SIC-2007 industries has been available since 2008. Annual employment figures by 4-digit industry to then map to the 106 industries used in the FAI input-output table. Employment is measured as total number of full-time equivalents (FTEs), with a part-time employee counted as 0.5 FTE. Employment can also be measured in hours, using the total hours usually worked per week.   

Firstly, we total the number of employees (excluding self-employed) by 4-digit industry within each quarterly dataset, using the LFS person weights to aggregate to the employment count into an estimate for the whole population. To obtain annual figures, we then average employment in each industry across the four quarters. Employment in 2020 is currently based only on quarter 1 data. This produces a panel of 4-digit industry employment over the period 2010-2020.  

### ASHE Earnings Data

We obtain annual earnings data from the statistical releases of the [Annual Survey of Hours and Earnings (ASHE) by the Office for National Statistics](https://www.ons.gov.uk/employmentandlabourmarket/peopleinwork/earningsandworkinghours/datasets/industry4digitsic2007ashetable16). The ASHE is an annual panel dataset which is a representative sample of 1% of employees registered for PAYE and completed by employers on behalf of employees using payroll records. It therefore contains reliable data on earnings, and among the summary datasets released by ONS are estimates of average annual earnings by sector. We use the data from the 2020 release. 

## Methodology 

### __Step (1) Recover the Multipliers__

### *Derive the model*

First transform the IO flow table (the $n$ * $n$ matrix of inter-industry purchases) into the __A__ matrix - known as the matrix of technical coefficients. This is obtained by dividing the elements of each column by total output. The elements of the matrix then represents the number of units of the sector in a given row that is required to make one unit of output in the given columns sector. 

Total output, in matrix notation, is then given by the following equation where $x$ is a vector of final demand, $Ay$ represents a vector of intermediate demand, and so $y$ is total demand which must equal total output:

$$
y = Ay + x
$$
$$
(I-A)y = x
$$

$$
y = (I-A)^{-1}x
$$

The matrix given by $(I-A)^{-1}$ is known as the *__Leontief Inverse Matrix__*. Using this matrix, the effect of any change in demand on total output can be recovered simply by re-writting the above expression in differences:

$$
\Delta y = (I-A)^{-1} \Delta x
$$

### *The Output Multiplier*

The multipliers are the coefficients which are used to translate changes in inputs into changes in output. Taking the IO equation:

$$
\Delta y = (I-A)^{-1} \Delta x
$$

An illustrative example using a simple three sector model shows the derivation of the sectoral output multipliers: 


$$
\left(\begin{array}{c} 
\Delta y_1 \\ 
\Delta y_2 \\
\Delta y_3
\end{array}\right)
=
\left(\begin{array}{ccc} 
a_{11} & a_{12} & a_{13} \\
a_{21} & a_{22} & a_{23} \\
a_{31} & a_{32} & a_{33} 
\end{array}\right)
\left(\begin{array}{c} 
\Delta x_1 \\ 
\Delta x_2 \\
\Delta x_3
\end{array}\right)
$$ 

The $a_{ij}$ are the Leontief matrix elements for industry $j$ purchases from industry $i$. $x_{j}$ are the changes in industry $j$ final demand - this can come from export demand, household demand, or government purchases.  We can express this as a system of linear equations:

$$
\Delta y_1 = (a_{11}*\Delta x_1) + (a_{12}*\Delta x_2) + (a_{13}*\Delta x_3) \\
\Delta y_2 = (a_{21}*\Delta x_1) + (a_{32}*\Delta x_2) + (a_{23}*\Delta x_3) \\
\Delta y_3 = (a_{31}*\Delta x_1) + (a_{32}*\Delta x_2) + (a_{33}*\Delta x_3)
$$

The total change in output is the sum of the $\Delta y$ vector. Summing up the three equations and collecting like terms gives:

$$
\Delta y = \Delta x_1(a_{11} + a_{21} + a_{31}) + \\ \Delta x_2(a_{12}+a_{22}+a_{32}) + \\\ \Delta x_3(a_{13}+a_{23}+a_{33})
$$
Defining $M_j = \sum^3_{n=1}a_{nj}$ i.e. $M_j$ for the $j$th sector is the sum of column $j$ in the leontief matrix, and represents the output multiplier - the change in total economy output for a one unit change in final demand for sector $j$:

$$
\Delta y = \Delta x_1M_1 + \Delta x_2M_2 + \Delta x_3M_3
$$

Which generalises simply to an input-output analysis with any number, $N$, of sectors:

$$
\Delta y = \sum^N_{n=1}\Delta x_nM_n 
$$

### *GVA and Employment Multipliers*

Gross value added (GVA) multipliers are based on the assumption that the proportion of total output which is comprised of GVA remains unchanged after a change in final demand. GVA is the value of output beyond the value of the inputs, and includes taxes, gross operating surplus, and compensation of employees. Defining for a given sector $n$ the GVA coefficient as $k$ we can calculate the change in GVA:

$$
k_n = \frac{GVA_n}{y_n} \\
$$
Rearranging:
$$
GVA_n = k_n y_n \\
$$
In matrix notation:
$$
\Delta GVA = k \Delta y 
$$
Substitute in the expression for $\Delta y$
$$
\Delta GVA = k \Delta y = \sum^N_{n=1}k_n\Delta x_nM_n 
$$

So the GVA multiplier for sector $n$ is simply the GVA proportion multiplied by the output multiplier - $k_n M_n$. 

A similar approach is taken to the creation of the employment multipliers. We calculate for each sector the employment/output ratio - the number of employees required to make £1 worth of output - and multiply this by the output multiplier. We therefore implicitly make the assumption that employment remains proportional to output, given any change in production.  

## Implementation of a 1% Reduction in Demand in R

### __Step (1) Produce Inputs__

The first step is to derive the inputs for the scenario to be modelled.

```{r inputs_stapm, eval = FALSE}

  inputs <- ReadInputs(data,
                       policy_effect_year = 2022,
                       n_years = 1)

```

The alternative is to apply percentage changes to total spending - in basic prices - to the alcohol and tobacco duty in the package data `tobacco_data`, `mesas_eng_wales`, and `mesas_scotland`. These data can be used to calculate the change in cash spending in basic prices for years 2016-2020 for both alcohol and tobacco. 

```{r inputs, echo = FALSE}

tobalciomodel::tob_alc_data[year == 2020, c("product","exp_bp")]

```

```{r inputs_agg, eval = TRUE}

  inputs <- GenInputs(prop_alc_on  = -0.01,
                      prop_alc_off = -0.01,
                      prop_tob_fm  = 0,
                      prop_tob_ryo = 0,
                      year = 2020)
```

Whether using `GenInputs` or `ReadInputs`, the output produced is a vector with the following figures - changes in off-trade alcohol expenditure, changes in on-trade alcohol expenditure, changes in tobacco expenditure (all in basic prices and millions of pounds), total change in tobacco and alcohol expenditure, and the net change in government revenue which results from changing consumption. 

```{r show_inputs, echo = FALSE}
  inputs

```


### __Step (2) Policy Parameters__

The next step is to set the policy parameters and apply these to the data created in step 2a. 

- Use the selected data from the previous stage
- Set alcohol policy parameters 
  - exogenous change
  - minimum unit price __(not yet added)__
  - tax changes __(not yet added)__
- apply a policy change to each of the 10 alcohol products. 
- choose whether or not to allow households to re-allocate spending. 
- choose whether or not government simultaneously changes its spending __(not yet added)__

The output of the function is the vector containing changes in final demand. An example of the function; simulate an exogenous change in demand - where consumption of each product of on-trade alcohol falls by 10%.

The above code equally splits the fall in demand for on-trade alcohol across the two on-trade sectors. Instead of a 0.5/0.5 split, we can simulate a distribution of changes in output by drawing a random proportion to allocate from a uniform distribution. 


### __Step (3) Calculate the Policy Impact__

Multiply the change in final demand vector calculated in step 2b by the appropriate column of multipliers obtained in step 1. 
