---
title: "Using `tobalciomodel` to Calculate Economic Impacts"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using 'tobalciomodel' to Calculate Economic Impacts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(tobalciomodel)
```

## Recover the Multipliers

- Draw on the package IO table data to construct output and other multipliers 
- Create the Leontief inverse matrices - Type I and Type II.
- Calculate sectoral output multipliers. 
- Calculate sectoral GVA coefficients (to adjust output multipliers into GVA multipliers)
- Caclulate sectoral GVA multipliers using the coefficients. 
- Repeat the GVA steps for employment __(not yet included, to be added)__

## Defining Input Parameters and Data

- Select the year of data to use.
- The choice of price elasticities to use.
- Whether the analysis is for Scotland or England & Wales

## Simulate the Policy 

- Use the selected data from the previous stage
- Set alcohol policy parameters - exogenous change/MUP/tax change - and select how these vary across the 10 alcohol products. - choose whether or not to allow households to re-allocate spending. 
- choose whether or not government simultaneously changes its spending __(not yet included, to be added)__

- function generates the change in final demand vector. 

## 
