---
title: "Using `tobalciomodel` to Calculate Economic Impacts"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using 'tobalciomodel' to Calculate Economic Impacts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width=6, fig.height=4.5
)
```

## __Step (0) Load Required Packages__

```{r setup, warning=FALSE, message=FALSE}
library(tobalciomodel)
library(data.table)
library(dplyr)
library(tidyverse)
```

## __Step (1) Recover the Multipliers__

- Draw on the package IO table data to construct output and other multipliers 
- Create the Leontief inverse matrices - Type I and Type II.
- Calculate sectoral output multipliers. 
- Calculate sectoral GVA coefficients (to adjust output multipliers into GVA multipliers)
- Caclulate sectoral GVA multipliers using the coefficients. 
- Repeat the GVA steps for employment __(not yet included, to be added)__

### *The Output Multiplier*

The multipliers are the coefficients which are used to translate changes in inputs into changes in output. Taking the IO equation:

$$
\Delta y = (I-A)^{-1} \Delta x
$$

An illustrative example using a simple three sector model shows the derivation of the sectoral output multipliers: 

$$
\left(\begin{array}{c} 
\Delta y_1 \\ 
\Delta y_2 \\
\Delta y_3
\end{array}\right)
=
\left(\begin{array}{ccc} 
a_{11} & a_{12} & a_{13} \\
a_{21} & a_{22} & a_{23} \\
a_{31} & a_{32} & a_{33} 
\end{array}\right)
\left(\begin{array}{c} 
\Delta x_1 \\ 
\Delta x_2 \\
\Delta x_3
\end{array}\right)
$$ 

The $a_{ij}$ are the Leontief matrix elements for industry $j$ purchases from industry $i$. $x_{j}$ are the changes in industry $j$ final demand - this can come from export demand, household demand, or government purchases.  We can express this as a system of linear equations:

$$
\Delta y_1 = (a_{11}*\Delta x_1) + (a_{12}*\Delta x_2) + (a_{13}*\Delta x_3) \\
\Delta y_2 = (a_{21}*\Delta x_1) + (a_{32}*\Delta x_2) + (a_{23}*\Delta x_3) \\
\Delta y_3 = (a_{31}*\Delta x_1) + (a_{32}*\Delta x_2) + (a_{33}*\Delta x_3)
$$

The total change in output is the sum of the $\Delta y$ vector. Summing up the three equations and collecting like terms gives:

$$
\Delta y = \Delta x_1(a_{11} + a_{21} + a_{31}) + \\ \Delta x_2(a_{12}+a_{22}+a_{32}) + \\\ \Delta x_3(a_{13}+a_{23}+a_{33})
$$
Defining $M_j = \sum^3_{n=1}a_{nj}$ i.e. $M_j$ for the $j$th sector is the sum of column $j$ in the leontief matrix, and represents the output multiplier - the change in total economy output for a one unit change in final demand for sector $j$:

$$
\Delta y = \Delta x_1M_1 + \Delta x_2M_2 + \Delta x_3M_3
$$

Which generalises simply to an input-output analysis with any number, $N$, of sectors:

$$
\Delta y = \sum^N_{n=1}\Delta x_nM_n 
$$

### *GVA and Employment Multipliers*

### *Implementation in R*

In R, recovery of the multipliers simply involves using the `multipliers()` function, which takes no arguments. The output is a data frame with 106 observations - 1 per sector - and 11 variables. The variables are the sector names and 5 multipliers (output,gross value added (total),tax,compensation of employees, and gross operating surplus), each calculated in the Type 1 and Type II scenarios.

```{r multipliers}
multipliers <- multipliers()
dim(multipliers)
names(multipliers)
head(multipliers[,c(2,3)])
```

## __Step (2a) Defining Input Parameters and Data__

- Select the year of data to use.
- The choice of price elasticities to use.
- Whether the analysis is for Scotland or England & Wales

```{r inputs}
inputs <- final_demand_inputs(yr = 2010,
                              scotland = FALSE,
                              elasticity = "meng14")

names(inputs)

head(inputs)
```
## __Step (2b) Simulate the Policy __

- Use the selected data from the previous stage
- Set alcohol policy parameters - exogenous change/MUP/tax change - and select how these vary across the 10 alcohol products. - choose whether or not to allow households to re-allocate spending. 
- choose whether or not government simultaneously changes its spending __(not yet included, to be added)__
- function generates the change in final demand vector. 

Simulate an exogenous change in demand - where consumption of each product of on-trade alcohol falls by 10%.

```{r policy}
policy_sim <- simulate_alcohol_policy(data = inputs,
                                      alc.policy = "exog",
                                      on.trade.ch = rep(-0.1,5),
                                      off.trade.ch = c(0,0,0,0,0),
                                      reallocate = FALSE,
                                      prob = FALSE)

names(policy_sim)

head(policy_sim)

policy_sim[61] # Wholesale/Retail (Alcohol)
policy_sim[69] # Accommodation (Alcohol)
policy_sim[71] # Food and Beverage (Alcohol)

```

The above code equally splits the fall in demand for on-trade alcohol across the two on-trade sectors. Instead of a 0.5/0.5 split, we can simulate a distribution of changes in output by drawing a random proportion to allocate from a uniform distribution. 


```{r simulation,include=FALSE}
### calculate the total change in output

output.effects <- sum(policy_sim*multipliers[,"output.multipliers.type1"])

### see the effect of the random splitting of the on-trade
set.seed(2020)
reps <- 1000
effects.vector <- rep(NA,reps)

for (i in 1:reps) {
  print(paste0("Simulation Iteration: ",i))
  policy_sim <- simulate_alcohol_policy(data = inputs,
                                        alc.policy = "exog",
                                        on.trade.ch = rep(-0.1,5),
                                        off.trade.ch = c(0,0,0,0,0),
                                        reallocate = FALSE,
                                        prob = TRUE)

  effects.vector[i] <- sum(policy_sim*multipliers[,"output.multipliers.type1"])
}
```

```{r sim_plot,echo=FALSE}

library(tidyverse)
data <- data.frame(effects.vector)

ggplot(data=data) +
  aes(x = effects.vector) +
  geom_density(fill="navy",alpha=0.4) +
  geom_vline(xintercept = output.effects) +
  theme_minimal() +
  labs(y = "Density",
       x = "Effect on Output (£bn)",
       title = "Effect of a 10% Fall in Consumption of On-Trade Alcohol Products",
       subtitle = "Distribution of Effects when Varying Split Across Two On-Trade Sectors") +
  scale_x_continuous(breaks = seq(-5.0,-3.0,0.1), minor_breaks = NULL)

```

## __Step (3) Calculate the Policy Impact__

Multiply the change in final demand vector calculated in step 2b by the appropriate column of multipliers obtained in step 1. 
